KIN: Method to infer relatedness using low coverage ancient DNA

### 2022-04-05 paper_overleaf/contamination.tex
Updates from Overleaf

\documentclass[12pt, letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage[super]{nth}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage[
singlelinecheck=false
]{caption}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{bbm}
\usepackage{graphicx}
\usepackage{csvsimple}
\usepackage[section]{placeins}
\usepackage{lineno}
\linenumbers

\title{KIN: Method to infer relatedness using low coverage ancient DNA}
\author{Divya Ratan Popli, Stéphane Peyrégne, Benjamin M. Peter}
\date{5 August 2021}
\linenumbers

\setlength{\parskip}{1em}
\setlength{\parindent}{0em}

\newcommand{\BZ}{\mathbf{Z}}
\newcommand{\BD}{\mathbf{D}}
\newcommand{\BN}{\mathbf{N}}
\newcommand{\BH}{\mathbf{H}}
\newcommand{\Btheta}{\pmb{\theta}}

\begin{document}
\nolinenumbers

\maketitle

\begin{abstract}

\noindent Knowledge of familial relationships is an important part of several research fields. Knowing genetic kinship of ancient specimens can lead to interesting insights into the culture and social hierarchy of ancient populations, and is also relevant for downstream genetic analyses. However, estimating relatedness from ancient DNA can be difficult since DNA sequences extracted from ancient remains may have low coverage, ascertainment bias, and contamination from modern human DNA. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child with as low as 0.05x coverage DNA data. The output can be further refined with another HMM that we developed to estimate ROH proportions. We focus on the application of KIN on ancient human DNA, but it is applicable to other diploid species as well. On a machine with 8 cores, the generation of input files from 17 bam files takes -- minutes, and running KIN software takes -- minutes.
\end{abstract}

\section{Introduction}

\subsection{Why study relatedness?}

Identifying related individuals is a common task in genetic studies. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential  step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.

In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity. ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}

\subsection{Approaches to estimate relatedness from high-coverage data}

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, each parent, for example, will share exactly one set of chromosomes IBD with their offspring. A grandparent will, on average share a quarter of their genome with a grand-child due to recombination. Along the genomes of a pair of diploid individuals, there are three IBD states possible at any given position: either no chromosomes are IBD, only one chromosome is IBD, or both chromosomes are IBD. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with proportions of 0.25,0.5,0.25, respectively (Fig \ref{fig0:schematic}). These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness $r$, which is defined as the proportion of the genome that is IBD. In the absence of inbreeding, this would be calculated as $r= k_1/2 + k_2$.

However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS \cite{vai_KINship_2020, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}. One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness \cite{roberson_visualization_2009}.

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings without and with ROH. In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases. p can take values of $p_0$, $p_1$ or $p_2$ which are expected proportion of differences in IBD states 0,1 and 2 respectively.  }
    \label{fig0:schematic}
\end{figure}

These methods utilize population allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}.  PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each Single Nucleotide Polymorphism (SNP) in the analyzed samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies in the samples (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}.

\begin{table}
\caption{\label{tab:Table 1}IBD sharing probabilities for different relations in absence of inbreeding}
\begin{tabular}{|c|c|c|c|}
    \hline
    Relatedness & $k_0$ & $k_1$ & $k_2$\\
    \hline
    Unrelated & 1 & 0 & 0\\
    \hline
    3rd Degree & 0.75 & 0.25 & 0\\
    \hline
    2nd Degree & 0.5 & 0.5 & 0\\
    \hline
    Siblings & 0.25 & 0.5 & 0.25\\
    \hline
    Parent-Child & 0 & 1 & 0\\
    \hline
    Identical/twins & 0 & 0 & 1\\
    \hline
\end{tabular}
\label{table1}
\end{table}

\subsection{Approaches that address problems with ancient DNA data}
One major issue with applying the above mentioned methods to ancient DNA data is that the sequence coverage is typically low, making it difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}. In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may be imprecise at lower coverages . Ancient DNA analyses often face additional challenges such as contamination from present-day DNA \cite{peyregne_authentict_2020}, absence of suitable reference panels, and ascertainment bias in the case of data generated from DNA capture.

In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population. It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen. In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding. Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.

Moreover, ancient DNA is commonly captured with a SNP array to enrich for sequences from the target species. Particularly methods based  on the fraction of sites in different IBS states are sensiteive to the ascertainment bias caused by this non-random selection of targted sites \cite{waples_allele_2019}.

Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ (REF). This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness. However, READ only infers the degree of relatedness, and only up to second degree.

\subsection{How our method works}
Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM)-based approach to estimate relatedness and IBD from low-coverage ancient DNA data. KIN can detect up to \nth{3}-degree relatives, and differentiates between siblings and parent-child. KIN is also able to take into account ROH, contamination and ascertainment. We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.

\section{New Approaches}\label{new_approaches}

To infer the case of relatedness, KIN fits one HMM for each pair of individual and for each relatedness case. The KIN-HMM infers IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account. The best-fit model is then assigned as the inferred relatedness. If the locations of ROH tracts are unknown, we provide another HMM (ROH-HMM) to coarsely estimate the location of ROH for samples with sufficient coverage ($\geq 0.1x$). Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.

\subsection{Model description}\label{method_overview}

The goal of KIN-HMM is to infer how two individuals are  relateded via the patterns of shared IBD-states along a pair of genomes. For this purpose,  we subdivide the genomes of a pair of individuals into $L$ large genomic windows (typically of size 10Mb), and infer the pattern of IBD-sharing for each relatedness case we consider (Unrelated, $5^{th}$ Degree, $4^{th}$ Degree, $3^{th}$ Degree, Grandparent-Grandchild, Avuncular, Half-siblings, Parent-Child, Siblings, and Identical). We then compare the likelihood between all models, and classify each pair to the model with the highest likelihood. We also return the most likely location of IBD tracts (using standard Viterbi algorithm), and the state posterior probablities $\BZ$.

The details of the likelihood computation are given in section \ref{ll}. As any HMM, KIN-HMM  require a set of n emission (section \ref{B}) and transition probability matrices. We assume the transition matrix for each relatedness case is fixed. Our emissions include a vector of parameters ($\delta$) that describes the variance in the data for each IBD state that we infer from the data (section ADD REF).

\subsection{Input of KIN-HMM}\label{hmm_input}
The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in section \ref{roh}.

$D_w$ is the number of sites that are different between a pair of individuals. For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated. The simplest approach is is to  randomly-sampled a read from each position ("pseudo-haploid" sequences) \cite{haak_massive_2015}. However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_{s=1}^{N_w} \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

\subsection{Log likelihood of the KIN-HMM}\label{ll}
The KIN-HMM uses $\BD$ and $\BN$ to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two shared chromosomes IBD, respectively. To take ROH into account, we define the variable $H_w \in$ ($0$, $1$, $2$) that designates that zero, one or both individuals are homozygous in window $w$. Since $H_w$ is unobserved, in practice we use the estimates from ROH-HMM $h_{wj} = P(H_w = j)$.

There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$. The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case. The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

Thus, the complete data likelihood for the HMM is

\begin{align}\label{eq:1}
\log P(\BD,\BZ|\BH, \theta, \delta) &= \log P(\BD|\BZ,\BH, \theta, \delta) + \log P(\BZ |\theta) \nonumber\\
&= \sum_w \log P(D_w|Z_w,H_w, N_w, \delta) + \sum_w \log P(Z_w |Z_{w-1},A) + \log P(Z_0|\pi)\text{.}
\end{align}

\subsection{Emission probability}\label{B}

Using this setup, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta, \delta)$ from equation \ref{eq:1} and optimize them for $\delta$. The simplest model is to assume that sites in each window are equally distributed and independent, in this case we could use the  binomial likelihood:
$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$

where, $p$ is the proportion of differences expected for a particular IBD and ROH state. If the two individuals are unrelated in a particular window (i.e. $Z_w = 0$), then the expected proportion of pairwise differences depends solely on the population history, and we denote this proportion with $p_0$. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}. Thus, $p(Z_w=i, H_w=0) = p_i$.

The proportion of differences between unrelated individuals $p_0$ is an important parameter. We follow READ \cite{kuhn_estimating_2018} and estimate $p_0$ as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated. We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$, because one, respectively two of the four compared chromosomes are identical in these cases.

The presence of inbreeding adds an additional complication, as the number of shared chromosome may  change. For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are between identical copies. However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero. Note however, that $p_0$ does not depend on the presence of ROH, since all comparisons are between unrelated chromosomes even if both individuals are homozygous at a particular locus.

Taken together, we can summarize $p$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
\begin{equation}\label{eq:2}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_0 & p_0 \\
p_1 & p_2 & p_3 \\
p_2 & p_3 & p_3
\end{array}\right]
\end{equation}

As explained above, $p_3 = 0$. However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$).

The effect of these considerations is that even though we have nine possible combinations of $Z_w$ and $H_w$ for each window, there are actually only four unique $p$-parameters $p_i$ with $i$ $\in$ (0,1,2,3).

\paragraph{Beta Binomial Model}
The Binomial Model laid out above assumes that sites are independent. However, due to genetic linkage, nearby sites will be correlated. As a result, we empirically find that the data often has considerably higher variance than would be expected from a binomial model (fig. S\ref{figS1:binom}). We take this into account by adding an overdispersion parameter $\delta$.Just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ depends on the number of chromosomes compared and so each of the  four $p_i$ has a corresponding $\delta_i$ parameter

Taken together, our emission probabilities are 
\begin{align}\label{eq:3}
P(D_{w}|Z_w,H_w,N_w, \delta) &\sim BB[D_w; p_i, \delta_i, N_w] \nonumber\\
&= \binom{N_w}{D_w}\frac{B(D_w+p_i \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})},
\end{align}
where $i$ is fully determined by the combination of $Z_w$ and $H_w$ (see eq 3).

This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$. We use this equation even if preprocessing steps result in non-integer $D_w$ and $N_w$, in which case we approximate the binomial coefficient using Gamma functions.

\subsection{Estimation of $\delta$}\label{delta}
We estimate the $\delta$-parameters using an Expectation-Maximization (EM) algorithm (REF). In an EM-algorithm, we iterate the Expecation and Maximiation Steps until the algorithm converges.

\paragraph{Initialization}
The value of $\delta_i$ is unknown to start with, and we set it to a random value between x and y.

\paragraph{Expectation step}
In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | D_w, H_w, \theta_i, \delta_i^{(t)})$ using the forward-backward algorithm, where $\delta_i^{(t)}$ is the current guess for $\delta$ for a given IBD state.

\paragraph{Maximization step}
The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and optionally the ROH state-probabilities $h_{w\omega}$ obtained from the ROH-HMM.

\begin{align}\label{eq:4}
\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta_{i \omega}^{(t-1)}, p_{i \omega})h_{w\omega}\gamma_{wi}
\end{align}

Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p_i$ and $\delta_i$, i.e.

\begin{align*}\label{eq:5}
g_{w0} &= \gamma_{w0}\\
g_{w1} &= \gamma_{w1} h_{w0}\\
g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
g_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
\end{align*}

So we can rewrite equation \ref{eq:4} as:
\begin{align}\label{eq:6}
\mathcal{C} &= \sum_{i=0}^3 \sum_{w=1}^L \log P(D_{w}|Z_w, H_w, N_w, \delta_i, p_i)g_{wi}
\end{align}

The cost function \ref{eq:6} has one independent term for each $i$ and so we can separate them and estimate each $\delta_i$ independently using the XXX algorithm implemented in \texttt{scipy.optimize} (REF scipy).

%\paragraph{Constraints on $\delta$ optimization}\label{delta_constraint}

Unconstrained optimization of the $\delta_i$ could result in some confounding of cases, We realized that different cases of relatedness have different number of IBD states possible. For example, siblings may have all three IBD states present while a Parent-Child has only $Z_w=1$. However, the parent-child model could fit data generated under the sibling model by assigning it a very high $\delta_1$, which would reduce performance  \ref{figS4:bndsbeta}. We avoid this problem by constraining the $\delta$ such that the beta distributions for the different cases overlap by at most one standard deviation.

\subsection{Model comparison}\label{model_comp}
To infer the most probable relatedness case, we run our model on all cases ( section \ref{method_overview}) and compare the resulting likelihoods. We output the relatedness corresponding to the maximum likelihood model, and the confidence as given by the log likelihood ratio between the two highest likelihood models. The log likelihood-ratio is commonly used in likelihood-ratio tests to compare nested models. However, in our cases, models are not nested, and so standard likelihood-ratio theory does not apply. For this reason, we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig \ref{fig3:cutoff}).

\paragraph{Grouping of cases}
Particularly for low-quality data, we may not be able to distinguish all cases. Thus, we group $4^{th}$ and $5^{th}$ degree relatives together with unrelated.  Similarly, we group  half-siblings, avuncular and grandparent-grandchild to  $2^{nd}$-degree relatives in the final results. We report the final pairwise classification in the following categories: Unrelated, Third Degree, Second Degree, Parent-Child, Siblings, Identical individuals.

\subsection{ROH estimation model}\label{roh}
Our HMM to detect ROH tracts works similar to the KIN-HMM described above, but we only consider one individual at a time, and only positions covered by at least two reads. For each site, we calculate the proportion of reads that carry different alleles, and sum them up in windows along the genome. We call the vector with the number of differences $\mathbf{\Delta}$, and the vector with the number of sites with at least two reads $\mathbf{N}$. Our model has two possible hidden states: homozygous state ($Y_w=3$), and non homozygous state ($Y-w=2$). As above, we collect the hidden states in a vector $\mathbf{Y} = (Y_1, \dots Y_w, \dots, Y_L)$. The complete data likelihood for the model in this case is then:

\begin{align}\label{eq:10}
    log P(\mathbf{\Delta},\mathbf{Y}|\Theta) &= log P(\mathbf{\Delta}|\mathbf{Y},\Theta) + log P(\mathbf{Y}|\Theta)\nonumber\\
 &= [\sum_{w} log P(\Delta_w|Y_w, \Theta) + \sum_{w} log P(Y_w|Y_w-1, \Theta)] + P(Y_0| \Theta)
\end{align}

Since the source of ROH may not be known,  we estimate both transitions and emissions. We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions at expected proportion of differences in a homozygous tract ($p_3$) and expected proportion of differences in a non-homozygous tract ($p_2$). The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window. The M-step for emissions in analogous to that in the KIN model (eq. \ref{eq:4}), and the optimization step here is done with the following cost function:

\begin{align}\label{eq:11}
\mathcal{C} = \sum_{w=1}^L \sum_{i=2}^3 P(\Delta_w|\delta_{i},p_{i}) \Gamma_{wi} ,
\end{align}
where $P(\Delta_w|\delta_{i},p_{i})$ is a beta-binomial probability with mean $p$ and overdispersion parameter $\eta$ similar to eq. \ref{eq:3}.

To estimate transitions, we initialize the transition matrix with 0.2 for the off-diagonal entries, and update it using the standard Baum-Welch update step \cite{baum_maximization_1970}.

Similar to the KIN-HMM, we avoid fitting issues by forcing all windows whose proportion of differences is larger than $p_2$ to be in the non-homozygous state (fig. S\ref{figS5:ROHforced}).

\subsection{Contamination correction}\label{contam}
Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}. Even levels of contamination may make samples look less similar, and thus distort relatedness analyses. This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations . To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.

We conceptualize contamination in the following way: If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read  will actually come from the desired individual with probability $1-C_i$. With probability $C_i$ the read comes from a contaminant. Thus, when comparing two reads, one drawn from individual $i$ and the other drawn from individual $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, both reads are endogenous and we can use them to infer relatedness. With probability $C_i(1-C_j) + C_j(1-C_i)$, one read is endogenous and the other is contaminant. Finally, with probability $C_iC_j$, both reads are contaminant. If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the $C_iC_j$ terms  can be ignored. Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.

We assume that the level of contamination is known (cite Stephane and Kay's review). We  estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$. To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$. Here, $c$ is a binary variable with value $0$ or $1$ denoting that the comparison at a site is between endogenous reads, or between endogenous and contaminant reads respectively. Similarly $d$ is a binary variable taking values 0 and 1 for seeing no difference, or a difference, when comparing the reads at a given site. We can further expand this probability:

@@

\section{Results}\label{results}

We predict different cases of relatedness with KIN, and test it's performance on simulated pedigrees. We use these simulations to find the limits of KIN, and compare it to READ and lcMLkin \cite{lipatov_maximum_2015}, two existing methods. Finally we apply KIN to published data and show that it can give reliable information about genetic kinship.

\paragraph{Motivating Example}
In Fig.\ref{fig1:ibd} we show the output of different relatedness cases when they are applied to data from a pair of siblings.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/IBDplot.png}
    \centering
    \caption{\textbf{Comparison of pairwise difference data  and inferred IBD fragments}:  The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings. Dashed lines represent $p_0$, $p_1$ and $p_2$ estimates. The second panel shows the true IBD state for each window. The remaining panels show the IBD states predicted by particular relatedness models. The log likelihood value for each model is shown on upper left corner of the panel. Light and shaded background represent distinct chromosomes.}
    \label{fig1:ibd}
\end{figure}

In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.

\paragraph{Critical Values}
 To investigate the limits of our method, we plotted the true-positive and false-positive rates when we use a particular difference in log-likelihood as a cutoff( Fig.\ref{figS10:cutoff},  Fig.\ref{fig4:Comparison_READ_KIN}).   Fig.\ref{figS10:cutoff} shows that for all relatedness cases except for $3^{rd}$-degree, the false positive rate is below $5\%$ even when simply selecting the model with the highest likelihood. We observe that for $3^{rd}$-degree relatives, using a cutoff of 1.0 brings down the false positive rate close to $5\%$ for all coverages except 0.05x. Thus, we  recommended to use a cutoff of 1 for all cases where ROH or contamination are not a concern.

\paragraph{Evaluating the ROH detection}
We next evaluate the performance of the ROH-HMM( Fig.\ref{fig2:ROH}). We observe that as coverage decreases, the raw heterozygosity estimates become more noisy. In the case without ROH (Fig \ref{fig2:ROH}A), we correctly infer that there is no ROH for all but two windows. We observe that the model predicts ROH probabilities well in all cases, even at a coverage of 0.1x \ref{tab:Table 2}.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/inkscape_finalImg/ROHplot_final.png}
    \centering
    \caption{Estimation of ROH probabilities along the genome in simulations. Top panel shows proportion of differences in a simulated individual in windows along the genome, and the bottom panel shows probability of seeing no ROH. (A) Simulation with no ascertainment, contamination, or ROH. (B) Simulation with ROH. (C) Simulation with ROH and ascertainment. (D) Simulation with ROH and contamination.}
    \label{fig2:ROH}
\end{figure}

\paragraph{Comparing KIN-HMM with READ}
In Figure Fig. \ref{fig4:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios. We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  except for $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages. We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this does not affect KIN performance too much.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/comparison_plot.png}
    \centering
    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH. Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated. 'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
    \label{fig4:Comparison_READ_KIN}
\end{figure}

\paragraph{Evaluating IBD prediction}
KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child. We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals. We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}). Fig. shows that we can predict IBD states for Unrelated $3^{rd}$ Degree, $2^nd Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling. On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
    \centering
    \caption{Evaluation of IBD estimation at different coverages. y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean.}
    \label{fig4:Comparison_READ_KIN}
\end{figure}

\subsubsection{Application to Chagyrskaya Neanderthals}
To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia. This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya . The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena (again cite Laurits paper). All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}. is dataset has all the issues that KIN deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). We also find that Chagyrskaya 06 and Chagyrskaya 14 are classified as parent-child with low confidence, but from the morphology they likely stem from the same individual. We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively). Further, we identify   Chagyrskaya07 and Chagyrskaya60 as 3rd-degree relatives.

Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and cannot distinguish between siblings and parent-child.

We compared our IBD estimates to lcMLKIN for all pairs for whom READ results (with s.d. >1) and KIN results ($\Delta LL$>1) match (fig. S\ref{figS6:Chagyrskaya_ibd}).  We find that lcMLKIN does create four different clusters for different relatedness. However, the clusters are strongly deviated from the expected values, likely due to the ROH.  Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $r \approx 0.6, k_0 \approx 0.33$.

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/kinplot.png}
    \centering
    \caption{Application of KIN on Neandertal specimens from Chagyrskaya. The color of a square represents the relatedness, while the number within denotes log likelihood ratio between the two maximum likelihood models.}
    \label{fig5:Chagyrskaya_KIN}
\end{figure}

>>>>

To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$. Here, $c$ is a binary variable with value $0$ or $1$ denoting that the comparison at a site is between endogenous reads, or between endogenous and contaminant reads respectively. Similarly $d$ is a binary variable taking values 0 and 1 for seeing no difference, or a difference, when comparing the reads at a given site. We can further expand this probability:

@@
